### 2011-08-30 trunk/papers/smectics/smectics.tex
Move papers directory one level down
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1454 40876441-3496-0410-a613-a2fca35b2ae2

\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{a}$}, K. Stratford\textit{$^{b}$} and M.E. Cates \textit{$^{a}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

@@

\noindent \normalsize{The abstract should be a single paragraph which summarises the content of the article. Any references in the abstract should be written out in full \textit{e.g.} [Surname \textit{et al., Journal Title}, 2000, \textbf{35}, 3523].}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

\footnotetext{\textit{$^{a}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
\footnotetext{\textit{$^{b}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}

@@

\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P. Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}

\section{Introduction}
In our everyday lives we are surrounded by complex fluids such as foams, colloidal dispersions, microemulsions, polymeric fluids, liquid crystals and many biofluids.
Their intricate microstructure gives rise to unusual rheological properties that are not found in simple Newtonian fluids and make them indispensable materials in numerous industrial applications.
Despite their abundance in the processing, energy or chemical industry complex fluids remain difficult to model as they normally consists of two or more separate phases which interact with each other and the solvent.
This in turn causes stresses and emerging flow properties like the viscosity to depend highly on the fow state and on the structure of the microphases. 
In order to capture their rich phenomenology and properties ab initio without using constitutive relations that only mimick their behaviour, computationally expensive approaches are necessary.
It is only now that we have the resources to simulate nonlinear flow of complex fluids on large scales\cite{Saksena09}.\\
In the present work we focus on a specific class of complex fluids, whose members share the common feature to undergo an order-disorder transition from an isotropic-disordered state at high temperature to a microphase separated state at low temperature.
They can be described with a Brazovskii free energy functional \cite{Brazovskii75} and comprise systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.
Especially the latter have attracted considerable interest because of their intricate morphology.
Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}. 
Moreover, in some regimes shear-induced transitions from the microphase separated state to ordered phases like the lamellar one occur \cite{Cates89,Koppi93,Fredrickson94}.
Lyotropic liquid crystals are known to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93}.
Shear-induced ordering takes also place in smectic liquid crystals \cite{Larson93,Hamley99}, making these systems an interesting object for rheological studies.\\
The purpose of this paper is twofold.
First we investigate the morphology and rheology of a two-dimensional lamellar system in steady shear flow.
This system has been studied extensively \cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}, but because of computational constraints almost all results were limited to two dimensions and smaller strains.
Here we quantify rheological properties and try to relate them to the defect pattern and the microstructure of the fluid.
In the second part of the analysis we compare the results in two and three dimensions in order to see how the dimensionality of the system influences the defect dynamics and phase kinetics.
\\ 
The governing equations in our model are a convection-diffusion equation for the scalar order parameter, which represents the density difference of the two phases and the Navier-Stokes equation with a stress tensor that contains extra contributions due to the presence of the lamellar interfaces.
This approach allows for stress-strain rate relations to emerge rather than just putting them in by means of consitutive equations.
In our approach the dynamical equations are solved using a combination of finite difference scheme for the convection-diffusion equation and the lattice Boltzmann method (LBM) \cite{Succi} for the Navier-Stokes equation.
The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries[], where it is superior to conventional methods of computational fluid dynamics.\\ 
This paper is organised as follows. In Sec.II we introduce the model, including its equilibrium properties, the full stress tensor and the dynamical equations.
We briefly sketch how we derive structural information from the order parameter and describe our defect quantification procedure, which measures the local alignment of the lamellar structure.
We present results of the two dimensional case in In Sec.III, discuss the morphology, average and local shear stress densities and first normal stress differences and their relation to the defect structure.
We also show structure factors for various flow states and flow curves.
In Sec. IV we compare our 2D-results with results in three dimensions. 
A summary with conclusions closes the paper in Sec.V.

>>>>

\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{b}$}, K. Stratford\textit{$^{c}$} and M.E. Cates \textit{$^{b}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

@@

\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, threfore in the nonlinear regime. In two dimensions we find that, according to the thermodynamic parameters, shear may either align the system and stabilise an almost regular lamellar phase, or induce the nucleation and proliferation of defects in the smectic layers. In this last case, defects may be progressively annealed by slowly decreasing the applied shear rate, so that there are marked memory effects and the rheology is history-dependent. Finally, selected simulations in three dimensions show shear-induced ordering rather than defect nucleation, thus pointing out to a dramatic effect of dimensionality on the physics of the system.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

\footnotetext{\textit{$^{a}$~University College of London, London, UK}}
\footnotetext{\textit{$^{b}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
\footnotetext{\textit{$^{c}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}

@@

%\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P. Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}

\section{Introduction}
Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurth etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid. Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibers constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character. Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in biosensing.

Rheology gives a powerful way to characterise macroscopic physical properties of a soft material such as its viscosity, elastic and loss moduli, which quantify, for instance, its rigidity, response to applied forces or stresses etc. While in a Newtonian, viscous fluid, shearing gives rise to a stress which is linearly proportional to the applied shear rate -- the slope being the viscosity -- in complex fluids flow is often non-Newtonian and such a simple linear relation no longer holds. Thus, viscoelastic complex fluids may be shear thickening or shear thinning, if their effective viscosity increases or decreases with applied shear. They may even spontaneously phase separate when flowing, giving rise to shear bands which are familiar in the rheology of liquid crystals in the vicinity of the isotropic-nematic transition, and in wormlike micelles.

The rheological properties of a non-Newtonian fluid are often linked to the microstrucure of the material. For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate. Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals. While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1. Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics. Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via preshearing, or not. In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample. An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state. This leads to increased dissipation and hence to a larger viscosity. A well-known example of permeation flow is the one which occurs e.g. in cholesterics which are sheared or pushed along the direction of the cholesteric helix. However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve. For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches. Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.

Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically. This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations. At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally. In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear. For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters. In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.

%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.

The governing equations in our model are a convection-diffusion equation for the scalar order parameter, which represents the density difference of the two phases and the Navier-Stokes equation with a stress tensor that contains extra contributions due to the presence of the lamellar interfaces.
In our approach the dynamical equations are solved using a combination of finite difference scheme for the convection-diffusion equation and the lattice Boltzmann method (LBM) \cite{Succi} for the Navier-Stokes equation.
The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.

Our main finding is that as anticipated above depending on system parameters we find that a smectic phase may lead to vastly different rheological responses. 
When thermodynamic forces are weaker with respect to viscous, flow-induced, ones, shear quickly orders the system into a lamellar phase, with only a few defects. The rheology in this regime is Newtonian, and the shear stress is linear in the applied shear rate. If, on the other hand, thermodynamic forces are larger, we observe that shear actually leads to the nucleation of new defects, and the rheology is highly nonlinear, with the shear rate proportional to $\dot{\gamma}^{\alpha}$, with $\alpha\simeq 0.87$, leading to a shear thinning behaviour. Such a behaviour is qualitatively reminiscent of the one found experimentally in XXX, where the numerical value of $\alpha$ was however different. In this regime we also find that the rheology is history dependent, as the defect density decreases for instance if the shear rate is ramped down gradually. Finally, intriguingly, dimensionality appears to play a key role: within our 3D simulations, we have indeed not been able to enter the nonlinear rheology regime with defect nucleation, and we always observe a shear-induced ordering into a lamellar phase. More research and parameter steering will be needed to make this observation more quantitative.
We note that sheared smectics have been studied extensively via simulations in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}, but because of computational constraints almost all results were limited to two dimensions and smaller strains than the ones considered here. Furthermore in this work we use Lees-Edwards boundary conditions which allow us to probe the true bulk rheology of the system, at variance with previous work which employed solid walls and different boundary conditions, which introduce finite size and boundary effects into the physics.

This paper is organised as follows. In Sec.II we introduce the model, including its equilibrium properties, the full stress tensor and the dynamical equations.
We briefly sketch how we derive structural information from the order parameter and describe our defect quantification procedure, which measures the local alignment of the lamellar structure.
We present results of the two dimensional case in In Sec.III, discuss the morphology, average and local shear stress densities and first normal stress differences and their relation to the defect structure.
We also show structure factors for various flow states and flow curves.
In Sec. IV we compare our 2D-results with results in three dimensions. 
A summary with conclusions closes the paper in Sec.V.
